---
title: "Final Project"
subtitle: "Research Article"
author: "David Sun, Elizabeth Morales, Jaehee Jeong,\nSarah Heuschele, Tyler Chun, Yolanda Jin"
date: "5/27/2021"
margins: 0.5cm
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r AWS Connection, message=FALSE, warning=FALSE, paged.print=FALSE}
projpath <- getwd()

if(!exists("xdbsock")) {
  xdbsock <- ""
  cat("\n", "Parameter 'xdbsock' not found, setting to empty string for general usage", "\n")
}

library(RMySQL)
library(knitr)
library(xtable)

drv <- dbDriver("MySQL")

PARTNER_MAS405_AWS_MY_DB_ADMIN_PW='18@@ROuserPW405fun'
PARTNER_MAS405_AWS_MY_DB_ADMIN_HOST='stats405db.cziy0a015m5d.us-east-1.rds.amazonaws.com'
PARTNER_MAS405_AWS_MY_DB_ADMIN_USER='ROuser'
PARTNER_MAS405_AWS_MY_DB_ADMIN_DBNAME='db1'
PARTNER_MAS405_AWS_MY_DB_ADMIN_PORT=3306

xdbuser <- PARTNER_MAS405_AWS_MY_DB_ADMIN_USER
xpw     <- PARTNER_MAS405_AWS_MY_DB_ADMIN_PW
xdbname <- PARTNER_MAS405_AWS_MY_DB_ADMIN_DBNAME
xdbhost <- PARTNER_MAS405_AWS_MY_DB_ADMIN_HOST
xdbport <- as.integer(PARTNER_MAS405_AWS_MY_DB_ADMIN_PORT)

con <- dbConnect(drv, user=xdbuser, password=xpw, dbname=xdbname, host=xdbhost, port=xdbport, unix.sock=xdbsock)
```

#### Tasks

## Abstract

Celiotomy, commonly refered to as “colic surgery” is a type of surgery around the abdomin of a horse. According to the American College of Vetinary Surgeons, colic surgery is used to address many issues that affect the longevity of a horse, most commonly used to address issues within the gastrointestinal tract within the intestines. The surgery is also used to adress a number of other lesians around the abdominal. 

## Introduction

## Method(s)
1. Data Collection
2. Exploratory Data Analysis 

## Data
Our data was sourced from the University of California, Irvine Machine Learning Repository. 

This specific dataset was donated August 6, 1989 by Will Taylor. It was created for use by Mary McLeish & Matt Cecile who belong to the Department of Computer Science at the University of Guelph in Guelph, Ontario, Canada.

The data has 299 observations and 27 parameters that range from continuous to discrete and nominal variables.
Below you can find an outline of the variables found within the dataset as well as their respective description.

**Variable Name**        **Description**
-----------------------  ------------------------------------------------------------------------
ID                       a row indicator which uniquely identifies each row
surgery                  was the horse treated with or without surgery                                        
age                      identify if the horse in an adult or a child (young identified as under 
                         6 months)     
hospital_number          the case number assigned to the horse (same number if a horse is 
                         treated twice)
rectal_temp              the temperature of the horse's rectum recorded in degrees Celsius
pulse                    a horse's heart rate measured in beats per minute
respiratory_rate         the number of breaths a horse takes per minute
temp_of_extremities      a subjective way to measure the condition of outermost circulation by 
                         the temperature
peripheral_pulse         a subjective way to measure the condition of outermost circulation by 
                         pulse
mucous_membrane          a record of the color of the mucous membranes as an indicator of 
                         circulation
capillary_refill_time    measuring the time to refill the capillaries 
pain                     a measure of the horse's pain level (not to be taken as a linear 
                         variable)
peristalsis              an indication of the activity in the horse's gut
abdominal_distention     measure of the amount of air/gas filling the horse's gut as an 
                         indicator of a health problem
nasogastric_tube         amount of gas coming out of the nasogastric tube
nasogastric_reflux       the amount of reflux within the nasogastric tube
nasogastric_reflux_ph    ph of the reflux within the nasogastric tube
rectal_exam_feces        the amount of feces found within the rectal
abdomen                  an indicator to how the abdomen (small and large intestine) is in 
                         relation to feces
packed_cell_volume       number of red cells in the blood by volume
total_protein            total amount of protein found in the blood
abdomo_appearance        the color of the fluid taken from the horse's abdominal cavity
abdomo_protein           total amount of protein found in the gut
outcome                  current status of the horse: alive, dead, or euthanized
surgical_lesion          could the lesion and/or problem be treated with surgery
lesion_1                 indicator of the site, type, subtype, specific code for lesion 1
lesion_2                 indicator of the site, type, subtype, specific code for lesion 2
lesion_3                 indicator of the site, type, subtype, specific code for lesion 3
cp_data                  indicates if there is pathology data present
-----------------------  ------------------------------------------------------------------------

## Conclusions

## Bibliography and References
Colic emergencies " large animal hospital " College of veterinary medicine " University of Florida. (n.d.). Retrieved May 28, 2021, from https://largeanimal.vethospitals.ufl.edu/hospital-services/surgery/colic/

Colic Surgery in Horses. (n.d.). Retrieved from https://www.acvs.org/surgical-procedures/colic-surgery-horses

