---
title: "Officials in the NHL who Worked in High Scoring Games"
subtitle: "Stats MAS405 Spring 2021:Assignment 4"

author: 
  - "Team Name : i need a \\</br\\>"
  - "David Sun, Elizabeth Morales, Jaehee Jeong"
  - "Sarah Heuschele, Tyler Chun, Yolanda Jin"
date: "5/9/2021"
fontsize: "12pt"
output: 
  pdf_document: 
    includes:
      in_header: header.tex
      toc: yes
      toc_depth: '3'
      geometry: left=10mm, right=10mm, top=25mm, bottom=25mm
---


```{r TOP, echo=FALSE, message=FALSE, eval=TRUE,warning=FALSE, results='hide'}
#usethis::edit_r_environ()

projpath <- getwd()
if(!exists("xdbsock")) {
    xdbsock <- ""
    cat("\n", "Parameter 'xdbsock' not found, setting to empty string for general usage", "\n")
}
library(RMySQL)
library(knitr)
library(xtable)
library(dplyr)
library(kableExtra)
drv <- dbDriver("MySQL")


xdbuser <- Sys.getenv("MAS405_AWS_DZES_DB_RO_USER") ### 
xpw     <- Sys.getenv("MAS405_AWS_DZES_DB_RO_PW") ### 
xdbname <- Sys.getenv("MAS405_AWS_DZES_DB_RO_DBNAME") ### 
xdbhost <- Sys.getenv("MAS405_AWS_DZES_DB_RO_HOST") ###
xdbport <- as.integer( Sys.getenv("MAS405_AWS_DZES_DB_RO_PORT") )

con <- dbConnect(drv, user=xdbuser, password=xpw, dbname=xdbname, host=xdbhost, port=xdbport, unix.sock=xdbsock)
```



```{r runQueryDate, echo=FALSE, message=FALSE,warning=FALSE, eval=TRUE, results='hide'}

##### get min and max date

qstr <- "SELECT MIN(date), MAX(date) FROM nhl_playersDate_1"

xx <- dbGetQuery(con, qstr)
xx

xMinDataDate <- xx[1, 1]
xMaxDataDate <- xx[1, 2]

xminDD <- format( strptime(xMinDataDate, "%Y%m%d"), "%Y-%m-%d")
xmaxDD <- format( strptime(xMaxDataDate, "%Y%m%d"), "%Y-%m-%d")

xpntgoal <- 3

```



## Introduction


<!--
%In the NHL, many officials will work together on a game. We were more curious in breaking apart the officials and looking at how many games each individual official worked.
-->

For Assignment 4, we wanted to look into which officials had referred in which players had scored more than 3 points within the game. We are analyzing NHL game and player
data from `r xminDD` to `r xmaxDD`.

```{r runQuery2, echo=FALSE, eval=TRUE,warning=FALSE, results='hide'}
final_query <- "WITH high_game AS
(
SELECT player.gameID, game.officials FROM nhl_playersDate_1 player
LEFT JOIN nhl_gameDate_1 game
ON player.gameID = game.gameID
WHERE player.goals > 3
)
select official_name,count(distinct gameID) gameCnt from (
(
SELECT substring_index(officials,';',1) official_name,
gameID FROM high_game
)
union all
(
SELECT
substring_index(substring_index(officials,';',2),';',-1) official_name
,gameID FROM high_game
)
union all
(
SELECT
substring_index(substring_index(officials,';',3),';',-1) official_name,
gameID FROM high_game
)
union all
(
SELECT
substring_index(officials,';',-1) official_name,gameID FROM high_game
)
) x
group by official_name
order by 2 desc, 1 asc"
output <- dbGetQuery(con, final_query)
output
```


```{r echo=FALSE,warning=FALSE}
output <- output %>%
  rename(Name = official_name, Count = gameCnt) %>%
  kable("latex",longtable=TRUE) %>%
  kable_styling(font_size = 7,position = "center",latex_options = c( "repeat_header"))  

output
```

For more details, we have a breakdown of the officials that partook in games where a player scored more than 3 points. The table below will show the date of the game, an abbreviation of the home and visiting teams, a list of officials during the game and finally a reference to the player who scored and that player’s time on the ice.

```{r runQuery1, echo=FALSE, warning=FALSE, eval=TRUE, message=FALSE}
query_1 = "select pl.date, gm.VT, gm.HT,gm.officials,pl.team,concat(pl.firstname,' ',pl.lastname) as First_Lastname,pl.timeonice TimeonIce,pl.goals from db1.nhl_playersDate_1 pl 
inner join db1.nhl_gameDate_1 gm on pl.gameid = gm.gameid
where goals > 3	;"

output_1 = dbGetQuery(con,query_1)

output_1[ , "VT" ]<-gsub("\xe9", "e", output_1[ , "VT" ])
output_1[ , "HT" ]<-gsub("\xe9", "e", output_1[ , "HT" ])
output_1[ , "officials" ]<-gsub("\xe9", "e", output_1[ , "officials" ])
output_1[ , "team" ]<-gsub("\xe9", "e", output_1[ , "team" ])

output_1[,"TimeonIce"]<-round(output_1[,"TimeonIce"],2)

captureUpper<-function(v){
s<-strsplit(output_1[,v], " ")
sapply(s, function(x){
  toupper(paste(substring(x, 1, 1), collapse = ""))
})
}

output_1[,"VT"]<-captureUpper('VT')
output_1[,"HT"]<-captureUpper('HT')
output_1[,"team"]<-captureUpper('team')


output_1 %>%
  kable("latex",booktabs=T) %>%
  kable_styling(font_size = 7,position = "center")


```


