---
title: "How Does Colic Affect a Horse's Life?"
subtitle: "Research Article - STAT 405: Final Project"
author: |
  | David Sun, Elizabeth Morales, Jaehee Jeong
  | Sarah Heuschele, Tyler Chun, Yolanda Jin
  |
  | Team Name: i need a \</br\>
date: "6/1/2021"
output:
  pdf_document:
    toc: yes
    toc_depth: 2
  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
margins: 0.5cm
fontsize: 12pt
line-height: 1.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r AWS Connection, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
projpath <- getwd()

if(!exists("xdbsock")) {
  xdbsock <- ""
  cat("\n", "Parameter 'xdbsock' not found, setting to empty string for general usage", "\n")
}

library(RMySQL)
library(knitr)
library(xtable)
library(dplyr)
library(ggplot2)
library(ggrepel)

drv <- dbDriver("MySQL")

xdbuser <- Sys.getenv("MAS405_AWS_YOLANDA_DB_RO_USER") ### 
xpw     <- Sys.getenv("MAS405_AWS_YOLANDA_DB_RO_PW") ### 
xdbname <- Sys.getenv("MAS405_AWS_YOLANDA_DB_RO_DBNAME") ### 
xdbhost <- Sys.getenv("MAS405_AWS_YOLANDA_DB_RO_HOST") ###
xdbport <- as.integer( Sys.getenv("MAS405_AWS_YOLANDA_DB_RO_PORT") )

con <- dbConnect(drv, user=xdbuser, password=xpw, dbname=xdbname, host=xdbhost, port=xdbport, unix.sock=xdbsock)

safe_colorblind_palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", 
                             "#44AA99", "#999933", "#882255", "#661100", "#6699CC", "#888888")
```


## Abstract

The purpose of this research article is to look at the presence of colic in horses and how certain courses of treatment (undergoing colic surgery or not) affects the longevity of the horse. The data gave us a look into 299 hospital cases of horses that were admitted due to poor health, and of those horses which lived, died, or were euthanized. Through exploratory analysis, we were able to look at horse outcomes given variables associated with colic and colic Surgery. Additionally, we explored the key characteristics of the horse’s condition to determine if the outcome could have been different had they received surgery, especially the horses that had symptoms related to colic Surgery.

## Introduction

Horses are regarded as powerful and courageous animal that are highly desired for riding, ranch work, jumping, racing, or dressage. As with any animal, there are always complications that may arise in regards to their health. This is especially true in horses. Horses can develop health problems at any stage in life and some are more susceptible to health problems than others. Our dataset focuses on horses who suffer from colic. colic is defined as having abdominal pains but in horses it is referred to problems in gastro-intestinal tract. Some common causes of colic are High grain based diets/Low forage diets, Moldy/Tainted feed, Abrupt change in feed, Parasite infestation , Lack of water consumption leading to impaction colics, Sand ingestion, Long term use of NSAIDS, Stress and Dental problems. About 10% of horses die from colic which is why its important to understand variables that can contribute to their status, outcome.

## Method(s)

1. Data collection

Our data was sourced from the University of California, Irvine Machine Learning Repository. This specific dataset was donated on August 6, 1989, by Will Taylor. It was created for use by Mary McLeish & Matt Cecile who belong to the Department of Computer Science at the University of Guelph in Guelph, Ontario, Canada.

2. Data cleaning

When we generated two-way tables, we did not include null values. However, a lot of observations have at least one null value of the variables, so we decided to keep all the observations.

3. Exploratory Data Analysis

To analyze the dataset, we performed Exploratory Data Analysis (EDA.) EDA is following under data dictionary.

## Data Dictionary
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
horse <- dbGetQuery(con, paste0("SELECT * FROM horse"))
```

The data has 299 observations and 27 parameters that range from continuous to discrete and nominal variables.
Below you can find an outline of the variables found within the dataset as well as their respective description.


**Variable Name**        **Description**
-----------------------  ------------------------------------------------------------------------
ID                       a row indicator which uniquely identifies each row
surgery                  was the horse treated with or without surgery                                        
age                      identify if the horse in an adult or a child (young identified as under 
                         6 months)     
hospital_number          the case number assigned to the horse (same number if a horse is 
                         treated twice)
rectal_temp              the temperature of the horse's rectum recorded in degrees Celsius
pulse                    a horse's heart rate measured in beats per minute
respiratory_rate         the number of breaths a horse takes per minute
temp_of_extremities      a subjective way to measure the condition of outermost circulation by 
                         the temperature
peripheral_pulse         a subjective way to measure the condition of outermost circulation by 
                         pulse
mucous_membrane          a record of the color of the mucous membranes as an indicator of 
                         circulation
capillary_refill_time    measuring the time to refill the capillaries 
pain                     a measure of the horse's pain level (not to be taken as a linear 
                         variable)
peristalsis              an indication of the activity in the horse's gut
abdominal_distention     measure of the amount of air/gas filling the horse's gut as an 
                         indicator of a health problem
nasogastric_tube         amount of gas coming out of the nasogastric tube
nasogastric_reflux       the amount of reflux within the nasogastric tube
nasogastric_reflux_ph    ph of the reflux within the nasogastric tube
rectal_exam_feces        the amount of feces found within the rectal
abdomen                  an indicator to how the abdomen (small and large intestine) is in 
                         relation to feces
packed_cell_volume       number of red cells in the blood by volume
total_protein            total amount of protein found in the blood
abdomo_appearance        the color of the fluid taken from the horse's abdominal cavity
abdomo_protein           total amount of protein found in the gut
outcome                  current status of the horse: alive, dead, or euthanized
surgical_lesion          could the lesion and/or problem be treated with surgery
lesion_1                 indicator of the site, type, subtype, specific code for lesion 1
lesion_2                 indicator of the site, type, subtype, specific code for lesion 2
lesion_3                 indicator of the site, type, subtype, specific code for lesion 3
cp_data                  indicates if there is pathology data present
-----------------------  ------------------------------------------------------------------------

## Horse Demographics

```{R echo=FALSE, message=FALSE, warning=FALSE}
horse %>%
  group_by(age,outcome) %>%
  summarise(n = n()) %>%
  ggplot(., aes(x=age,fill=outcome,y = n)) + 
    geom_bar(stat = "identity",position=position_dodge()) +
    geom_text(aes(label = n), size = 3, position=position_dodge(width=0.9), vjust=-0.25)+
  labs(x = "Age of Horse", y = "Count of Horses", fill = "Outcome", title = "Horses by Age and Outcome") +
  scale_fill_manual(values=safe_colorblind_palette[1:3])+
theme(axis.title.x = element_text(size = 9), axis.title.y = element_text(size = 9), plot.title = element_text(hjust = 0.5, size = 11))
```

This graph shows number of horses by age groups and outcome which indicates horses' life status. The number of young horses (24) is very small so that it might not be enough to talk about the pattern. The bar graph shows that adult horses have higher lived proportion than young horses. 
Out of two hundred and seventy-five adult horses, one hundred and sixty-six (60%) horses have lived after treatment, whereas only 50% of young horses lived.

## Hospital Demographics 

There were 16 horses that has multiple treatments. Of those horses, we can see a breakdown of how many lived, died, or were euthanized. 

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
tb1<-horse %>%
  group_by(hospital_number) %>%
  summarise(n = n_distinct(ID)) %>%
  filter(n > 1) %>%
  left_join(horse, by = c('hospital_number' = 'hospital_number')) %>%
  group_by(outcome) %>%
  summarise(n = n_distinct(hospital_number))
knitr::kable(as.data.frame(tb1))
```

Comparing the horses that came back to the hospital numerous times to the horses that only visited once. We can see that there is a larger gap in the proportion of horses that need to be euthanized once returning to the hospital. The proportion of horses that returned to the hospital and were either euthanized or died (50%) is much larger than the proportion of horses that only went to the hospital once (39%). This helps to show that disregarding all other factors and demographics, if a horse was to return to the hospital, they had a 20% higher chance of being euthanize or dying. 

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
data <- horse %>%
  group_by(hospital_number) %>%
  summarise(n = n_distinct(ID)) %>%
  left_join(horse, by = c('hospital_number' = 'hospital_number')) %>%
  mutate(grouping = case_when(
    n > 1 ~ "many",
    TRUE ~ "once")) %>%
  group_by(grouping, outcome) %>%
  summarise(n = n_distinct(ID)) %>%
  mutate(class = paste0(grouping," ",outcome)) %>%
  mutate(class2 = case_when(
    class == "many died" ~ "Multiple Visits - Died",
    class == "many euthanized" ~ "Multiple Visits - Euthanized",
    class == "many lived" ~ "Multiple Visits - Lived",
    
    class == "once died" ~ "One Visits - Died",
    class == "once euthanized" ~ "One Visits - Euthanized",
    class == "once lived" ~ "One Visits - Lived"
  )) %>%
  mutate(class3 = case_when(
    class == "many died" ~ 1,
    class == "many euthanized" ~ 3,
    class == "many lived" ~ 5,
    
    class == "once died" ~ 2,
    class == "once euthanized" ~ 4,
    class == "once lived" ~ 6
  ))

ggplot(data, aes(x="", y=n, fill=reorder(class2, -class3))) +
  geom_bar(stat="identity", width=1, color="black") +
  #geom_label_repel(aes(label = class2,y =n),
  #                 size = 4, min.segment.length = 0, show.legend = FALSE, box.padding = 2,         #                  position= "stack",segment.linetype = 6,
  #                 segment.curvature = -1e-20) +
  geom_text(aes(label = paste(round(n / sum(n) * 100, 1), "%"), x = 1.3), color="white",
            position = position_stack(vjust = 0.5)) +
  coord_polar("y", start=0) +
  labs(title = "Outcome of Horse by Visits to Hospital") +
  scale_fill_manual(values = safe_colorblind_palette[1:6],
                    guide_legend(title = "Visits and Outcome")) +
  theme(plot.title = element_text(hjust = 0.5, size = 11))+
  xlab(' ')+ylab(' ')

```

## Physical Attributes

A normal rectal temperature for a horse sits around 37.8. A temperature above can come from an infection while a low temperature may be due to shock in the body. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=horse,aes(x=rectal_temp,fill=outcome)) + 
  geom_histogram(bins=10,color="black") + labs(title = "Horses Rectal Temp Distribution Across Outcomes", fill = "Outcome", x = "Rectal Temperature",y="Count") +
  scale_fill_manual(values=safe_colorblind_palette[1:3])+
  theme(axis.title.x = element_text(size = 9), axis.title.y = element_text(size = 9), plot.title = element_text(hjust = 0.5, size = 11))

```

The average temperatures are similar between the three outcomes; died, euthanized, and lived, but  temperatures of the horses which died are slightly higher with more cases (16%) ranging from 39 to 41 degree, and euthanized horses’ temperaturesare lower than the other two in the  range. There are higher variation for died and euthanized horses. The distribution for lived horses had more stable rectal temperatures than the other two groups.

## Circulation Attributes

Cold and cool temperatures are associated with possible shock of the horse, while hotter extremities (in this case denoted as 'Warm') are associated with elevated rectal temperatures. Both cases signal that there is an underlying health issue with the horse. 

```{R echo=FALSE, message=FALSE, warning=FALSE}
library(gridExtra)
bg1<-horse %>%
  filter(temp_of_extremities != "NA") %>%
  group_by(outcome, temp_of_extremities) %>%
  summarise(n = n_distinct(ID)) %>%
  ggplot(., aes(fill=temp_of_extremities, y=n, x=outcome)) + 
  geom_bar(position="fill", stat="identity") +
  labs(title = 'Proportion of Extremity\n Temperatures by Outcome',fill = 'Temperature',
       x ="Outcome", y = 'Proportion of Temperatures') + 
  theme(legend.position = "bottom", legend.key.width=unit(0.1,'cm'))+
  scale_fill_manual(values=safe_colorblind_palette[1:4])+
  theme(axis.title.x = element_text(size = 9), axis.title.y = element_text(size = 9), plot.title = element_text(hjust = 0.5, size = 11), plot.margin = unit(c(0.2,0.2,0.2,0.2), "cm"))


bg2<-horse %>%
  filter(peripheral_pulse != "NA") %>%
  mutate(grouping  = case_when(
    peripheral_pulse == 'increased' ~ 'normal',
    peripheral_pulse == 'normal' ~ 'normal',
    peripheral_pulse == 'absent' ~ 'problematic',
    peripheral_pulse == 'reduced' ~ 'problematic'
  )) %>%
  group_by(outcome, grouping) %>%
  summarise(n = n_distinct(ID)) %>%
  ggplot(., aes(fill=grouping, y=n, x=outcome)) + 
  geom_bar(position="fill", stat="identity") +
  labs(title = 'Proportion of Good and Bad \n Peripheral Pulse',fill = 'Indicator', x = "Outcome", y = 'Proportion of Pulse') + 
  theme(legend.position = "bottom", legend.key.width=unit(0.3,'cm')) +
  scale_fill_manual(values=safe_colorblind_palette[1:2])+
  theme(axis.title.x = element_text(size = 9), axis.title.y = element_text(size = 9), plot.title = element_text(hjust = 0.5, size = 11), plot.margin = unit(c(0.2,0.2,0.2,0.2), "cm"))



# bg3<-horse %>%
#   filter(mucous_membrane != "NA") %>%
#   group_by(mucous_membrane, outcome) %>%
#   summarise(n = n_distinct(ID)) %>%
#   ggplot(., aes(x=outcome, y=n, fill=mucous_membrane)) + 
#   geom_bar(position="fill", stat="identity") +
#   labs(title = 'Proportion of Mucous Membrane',fill = 'Indicator', x = "Outcome", y = 'Proportion of Mucous Membrane') + 
#   theme(legend.position = "bottom", legend.key.width=unit(0.3,'cm'))
bg3<-horse %>%
  filter(mucous_membrane != "NA") %>%
  mutate(grouping = case_when(
    mucous_membrane == 'normal_pink' ~ 'normal',
    mucous_membrane == 'bright_pink' ~ 'normal',
    mucous_membrane == 'pale_cyanotic' ~ 'problematic',
    mucous_membrane == 'pale_pink' ~ 'problematic',
    mucous_membrane == 'dark_cyanotic' ~ 'problematic',
    mucous_membrane == 'bright_red' ~ 'problematic'
  )) %>%
  group_by(outcome, grouping) %>%
  summarise(n = n_distinct(ID)) %>%
  ggplot(., aes(x=outcome, y=n, fill=grouping)) + 
  geom_bar(position="fill", stat="identity") +
  labs(title = 'Proportion of Mucous Membrane',fill = 'Indicator', x = "Outcome", y = 'Proportion of Mucous Membrane') + 
  theme(legend.position = "bottom", legend.key.width=unit(0.3,'cm')) +
  scale_fill_manual(values=safe_colorblind_palette[1:4])+
  theme(axis.title.x = element_text(size = 9), axis.title.y = element_text(size = 9), plot.title = element_text(hjust = 0.5, size = 11), plot.margin = unit(c(0.2,0.2,0.2,0.2), "cm"))

#plist <- list(bg1,bg2,bg3,bg3)
#grid.arrange(grobs = plist, ncol = 2) 
#wrap_plots(plist,heights = 10)
#ggarrange(plotlist = plots, ncol=2, nrow = 2)

grid.arrange(bg1,bg2 + scale_fill_manual(values=safe_colorblind_palette[1:2]),ncol=2)
grid.arrange(bg3 + scale_fill_manual(values=safe_colorblind_palette[1:2]),ncol=2)
```

Peripheral Pulse is a measure of how fast blood, in the outermost parts of the horse, is moving away from the heart way and is being used to measure the horse's circulation. A normal or increased peripheral pulse is considered normal, however, a reduced or absent pulse in an indicator of poor circulation.


The Mucos Membrane color can also be an indicator of poor circulation and lead to an earlier admission into the hospital. The following table denotes the meaning behind the colors;



**Color**                 **Meaning**
-----------------------  ------------------------------------------------------------------------
Normal Pink               indicate a normal or slightly increased circulation
Bright Pink               indicate a normal or slightly increased circulation
Pale Pink                 occurs in early stages of shock
Pale Cyanotic             indicates of a serious circulatory issue
Bright Red/Injected       indicates septicemia
Dark Cyanotic             indicates of a serious circulatory issue
-----------------------  ------------------------------------------------------------------------

capillary refill time
The longer the refill, the poorer the circulation

## Pain Attributes

```{r TwoWayTable_Pain_Outcome, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Display two-way table for Pain and Outcome
library(pander)
query9 <- "SELECT * FROM db1.horse;"
result9 <- dbGetQuery(con, query9)
twoWay_Pain_Outcome <- pander(addmargins(table(result9$pain, result9$outcome)))
twoWay_Pain_Outcome
```

This two way table states that the more pain a horse has, it is less likely to be "lived" than "died" or "euthanized". For example, when horses feel extreme pain, there are only 1/3 of horses to be "lived" when the other 2/3 were either "died" or "euthanized". However, when horses alert status, over 90% of horses to be "lived".

## Outcome Analysis

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
horse %>%
  group_by(surgical_lesion, outcome) %>%
  summarise(n = n()) %>%
  ggplot(., aes(x=surgical_lesion,fill=outcome,y = n)) + 
  geom_bar(stat = "identity",position=position_dodge()) +
  labs(x = "Indicator if Surgery was Performed", y = "Count of Horses", fill = "Outcome", title = "Horses Outcome vs Surgery Performed")+
  scale_fill_manual(values=safe_colorblind_palette[1:3])+
  geom_text(aes(label = n), size = 3, position=position_dodge(width=0.9), vjust=-0.25)+
  theme(axis.title.x = element_text(size = 9), axis.title.y = element_text(size = 9), plot.title = element_text(hjust = 0.5, size = 11))
```

This table provides some input to the most common lesions found on horses and an analysis to compare the horses demographics of age, outcome and if they had surgery in respect to the place of the lesion. 

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
table <- horse %>%
  group_by(lesion_1) %>%
  summarise(n = n_distinct(ID)) %>%
  filter(n > mean(n)) %>%
  left_join(horse, by = c('lesion_1' = 'lesion_1')) %>%
  select(age, outcome, surgery, lesion_1, ID) %>%
  ftable(., row.vars=c("age", "outcome"), col.vars=c("surgery", "lesion_1"))

df = as.data.frame(table) %>% 
  arrange(desc(Freq))

knitr::kable(df[1:10,])


```

## Conclusions

While it is common for horses to develop health issues, such as colic, each horse may develop different symptoms and respond differently to various treatments. Specifically, because our dataset included both adult and young horses, we expect them to share certain effects of colic but respond differently as their immune systems are at different stages. From the horse demographics, we have found that even though there is a very small number of young horses, there is a higher “lived” adult horse proportion than young horses. This might be due to adult horses having more robust health system to recover from surgeries and illness. As with any patient, complications may arise post diagnosis or surgery and they may need to revisit the Hospital for further treatment. In our dataset, Horse’s hospital demographics show that if a horse was to return to the hospital, they had a 20% higher chance of being euthanized or dying. Temperatures measurements, including rectal and peripheral pulses are strong indicators of a horse's stability in response to the infection or shock to its immune system. Those who lived have three times higher proportions (around 60%) of normal and warm temperatures than those who died or euthanized (18~20%). About 70% of lived horses had normal peripheral pulse and about 70% of died or euthanized horses had problematic peripheral pulses. The distribution for lived horses had more stable rectal temperatures than the other two groups. About 60% of the lived horses had normal circulation when about 20~25% of the died or euthanized horses had normal circulation. According to the dataset, more severe conditions indicated by higher pain level will result in lower survival rate. About 70% of the horses without surgery survived when horses who had surgery, around 53% lived, and 32% died. As a result, we can utilize this analysis of horses' conditions to better predict their survival rate on the various treatments.

## Bibliography and References

Colic emergencies " large animal hospital " College of veterinary medicine " University of Florida. (n.d.). Retrieved May 28, 2021, from https://largeanimal.vethospitals.ufl.edu/hospital-services/surgery/colic/

Colic Surgery in Horses. (n.d.). Retrieved from https://www.acvs.org/surgical-procedures/colic-surgery-horses
